---
title: "Probability Distributions and Standard Error"
author: "Jeffrey Heinz"
date: "2/24/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

R has built-in functions for different kinds of probability distributions. These often come in families. (Replace *name* with the keyword associated with the distribution)

* rname: generates a random sample according to the distribution.
* dname: returns the value of the probability density function
* pname: returns the value of the cumulative density function
* qname: returns the value of the inverse cumulative density function
  
We won't study the inverse cumlative density function here.

Let's see some examples.

```{r seed}
library(tidyverse)
set.seed(101)
```

## Normal

```{r normal}
data <- rnorm(1000,mean=100,sd=10)
data %>% as_tibble() %>% ggplot(aes(x=data)) + geom_histogram(bins=20)
```


Let's look at the probability density.

```{r normal df}
x <- 1:200
x %>% as_tibble() %>% ggplot(aes(x)) + 
  stat_function(fun = function(x) dnorm(x, mean = 100, sd = 25))
```

Let's look at the cumulative probability density.

```{r normal cdf}
x <- 1:200
x %>% as_tibble() %>% ggplot(aes(x)) + 
  stat_function(fun = function(x) pnorm(x, mean = 100, sd = 25))
```
## Chi Square

```{r Chi Square}
data <- rchisq(1000,df=5)
data %>% as_tibble() %>% ggplot(aes(x=data)) + geom_histogram()
```
Let's look at the probability density.

```{r chisq df}
x <- 0:20
x %>% as_tibble() %>% ggplot(aes(x)) + 
  stat_function(fun = function(x) dchisq(x, df=5))
```

Let's look at the cumulative probability density.

```{r chisq cdf}
x <- 0:20
x %>% as_tibble() %>% ggplot(aes(x)) + 
  stat_function(fun = function(x) pchisq(x, df=5))
```

## T distribution

```{r student t }
data <- rt(1000,df=10,ncp = 10)
data %>% as_tibble() %>% ggplot(aes(x=data)) + geom_histogram()
```
Let's look at the probability density.

```{r student df}
x <- 0:20
x %>% as_tibble() %>% ggplot(aes(x)) + 
  stat_function(fun = function(x) dt(x, df=10,ncp=10))
```

Let's look at the cumulative probability density.

```{r student cdf}
x <- 0:20
x %>% as_tibble() %>% ggplot(aes(x)) + 
  stat_function(fun = function(x) pchisq(x, df=10,ncp=10))
```

## Bernoulli (discrete)

```{r Bernoulli}
data <- rbernoulli(1000,p=.5)
data %>% as_tibble() %>% ggplot(aes(x=data)) + geom_bar()
```



## Possion (discrete)

```{r Poisson}
data <- rpois(1000,lambda = 5)
data %>% as_tibble() %>% ggplot(aes(x=data)) + geom_bar()
```





