---
title: "R Notebook on Regression (based on Winter 2020)"
output: html_notebook
---

First we load the tidyverse and read the "fishCatchData.csv" as a tibble.

```{r}
library(broom)
library(tidyverse)
fish <- read_csv('fishCatchData.csv')
tibble(fish)
mean(fish$Weight_g)
```

Let's plot it!

```{r}
plot(fish$Length_cm, fish$Weight_g, pch=19)   # (plot(x,y) puts x on x-axis and y on y-axis)
```
Next we make our linear model, which essentially just finds the coefficients that minimize the residuals.
```{r}
model <- lm(Weight_g ~ Length_cm, fish)
coef(model)
```

The summary function summarizes key aspects of the model.
```{r}
summary(model)
```

Finally, let's plot the model using Winter 2020's plotting recommendations.
```{r}
fish %>% ggplot(aes(x = Length_cm, y = Weight_g)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()
```

Now we add columns to the fish tibble to get our centered Length values and normalized 'z' length values.
```{r}
fish <- mutate(fish, LengthCenter = Length_cm - mean(Length_cm), LengthZ = LengthCenter / sd(LengthCenter))
tibble(fish)
```
Note in the future the R function scale(x) will compute x/sd(x).


Next we create linear models with the centered and z predictors

```{r}
model_c <- lm (Weight_g ~ LengthCenter, fish)
model_z <- lm (Weight_g ~ LengthZ, fish)
```

When comparing model and model_c, the intercepts are different but the slopes are the same. This is because only difference is centering (shifting the line over).
```{r}
coef(model)
coef(model_c)
```

When comparing model_c and model_z, the intercepts are the same but the slopes are different. This is because the units have changed from centimeters to standard deviations.

```{r}
coef(model_c)
coef(model_z)
```
However, the quantitative measure of the variance each models captures is identical. They are basically the same under these linear transformations.
```{r}
glance(model)$r.squared
glance(model_c)$r.squared
glance(model_z)$r.squared
```

Finally, let's plot these models using Winter 2020's plotting recommendations.

```{r}
fish %>% ggplot(aes(x = LengthCenter, y = Weight_g)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()
fish %>% ggplot(aes(x = LengthZ, y = Weight_g)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()

```


Let's do a nonlinear transformation on the weight and see what that does.
```{r}
fish <- mutate(fish, LogWeight = log(Weight_g,10))
tibble(fish)
```

```{r}
plot(fish$Length_cm, fish$LogWeight, pch=19)   # (plot(x,y) puts x on x-axis and y on y-axis)
```
Let's try to predict the log weight from the length.

```{r}
model_log_z <- lm (LogWeight ~ LengthZ, fish)
fish %>% ggplot(aes(x = LengthZ, y = LogWeight)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()
glance(model_log_z)$r.squared
```


We can check for correlation by cestimating how much change in y standard units results in a corresponding change in x standard units. The resulting slope has a special name, it is Pearson's r, the "correlation coefficient".
```{r}
fish <- mutate(fish, WeightZ = scale(Weight_g))
tibble(fish)
```
Use the cor function to find the correlation. It is always between -1 and 1 with values close to 0 indicating little to no correlation.

```{r}
cor(fish$WeightZ,fish$LengthZ)
```

