---
title: "R Notebook on Regression (based on Winter 2020)"
output: html_notebook
---

First we load the tidyverse and read the "fishCatchData.csv" as a tibble.

```{r}
library(broom)
library(tidyverse)
fish <- read_csv('fishCatchData.csv')
tibble(fish)
mean(fish$Length_cm)
```

Let's plot it!

```{r}
plot(fish$Length_cm, fish$Weight_g, pch=19)   # (plot(x,y) puts x on x-axis and y on y-axis)
```


Next we make our linear model, which essentially just finds the coefficients that minimize the residuals.
weight = intercept + x*length

```{r}
model <- lm(Weight_g ~ Length_cm, fish)
coef(model)
```

The summary function summarizes key aspects of the model.
```{r}
summary(model)
```

Finally, let's plot the model using Winter 2020's plotting recommendations.
```{r}
fish %>% ggplot(aes(x = Length_cm, y = Weight_g)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()
```

Now we add columns to the fish tibble to get our centered Length values and normalized 'z' length values.
```{r}
fish <- mutate(fish, LengthCenter = Length_cm - mean(Length_cm), LengthZ = LengthCenter / sd(LengthCenter))
tibble(fish)
```
Note in the future the R function scale(x) will compute x/sd(x).


```{r}
fish %>% ggplot(aes(x = LengthCenter, y = Weight_g)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()
```



Next we create linear models with the centered and z predictors

```{r}
model_c <- lm (Weight_g ~ LengthCenter, fish)
model_z <- lm (Weight_g ~ LengthZ, fish)
```

When comparing model and model_c, the intercepts are different but the slopes are the same. This is because only difference is centering (shifting the line over).
```{r}
coef(model)
coef(model_c)
```

When comparing model_c and model_z, the intercepts are the same but the slopes are different. This is because the units have changed from centimeters to standard deviations.

```{r}
coef(model_c)
coef(model_z)
```
However, the quantitative measure of the variance each models captures is identical. They are basically the same under these linear transformations.
```{r}
glance(model)$r.squared
glance(model_c)$r.squared
glance(model_z)$r.squared
```

Finally, let's plot these models using Winter 2020's plotting recommendations.

```{r}
fish %>% ggplot(aes(x = LengthZ, y = Weight_g)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()
```


Let's do a nonlinear transformation on the weight and see what that does.
```{r}
fish <- mutate(fish, LogWeight = log(Weight_g,10))
tibble(fish)
```

```{r}
plot(fish$Length_cm, fish$LogWeight, pch=19)   # (plot(x,y) puts x on x-axis and y on y-axis)
```
Let's try to predict the log weight from the length.

```{r}
model_log_z <- lm (LogWeight ~ LengthZ, fish)
fish %>% ggplot(aes(x = LengthZ, y = LogWeight)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()
glance(model_log_z)$r.squared
```


We can check for correlation by estimating how much change in y standard units results in a corresponding change in x standard units. The resulting slope has a special name, it is Pearson's r, the "correlation coefficient".
```{r}
fish <- mutate(fish, WeightZ = scale(Weight_g))
tibble(fish)
```
Use the cor function to find the correlation. It is always between -1 and 1 with values close to 0 indicating little to no correlation.

```{r}
cor(fish$WeightZ,fish$LengthZ)
```


Next we want to see if the type of fish canhelp us predict the weight. We need to make the "Name" a *factor*. 
```{r}
fish <- mutate(fish, Name=factor(Name))
tibble(fish)
```


We can also use the function 'str' to see the structure of the tibble. Note also that 'summary' reports the factor column differently than before.
```{r}
str(fish)
summary(fish)
```

```{r}
table(fish$Name)
```

We can also group our data by the levels in the factors. Remember, in statistics speak, the factors contain levels.
```{r}
fish %>% group_by(Name) %>% summarize(M=mean(Weight_g), SD = sd(Weight_g))
```
The reference level is the first factor in alphanumeric order. In our case, this is 'bream'. We can change this later.


Don't forget that the whiskers in the boxplots correspond to the maximum of Q3 + 1.5 * IQR and the minimum of Q1 − 1.5 * IQR 
where IQR is the interquartile range. The middle line is the median, not the mean! Boxplots...

```{r}
fish %>% ggplot(aes(x = Name, y = Weight_g, fill = Name)) +
geom_boxplot() + theme_minimal() +
scale_fill_brewer(palette = 'PuOr')
```

We can now see how well we can predict the weight if we know the name (and only the name).
```{r}
model_name <- lm (Weight_g ~ Name , fish)
summary(model_name)
```
What does the above mean? The bream is the origin of a 6 dimensional space. There is Parkki dimension, Perch dimension, Pike dimension and so on. The estimates are the coefficients of the linear equation.
weight_g = 626 -471.18Parkki -243.76Perch +92.71Pike -465.95Roach -614.82Smelt -95.00Whitefish

The values of Parkki, Perch etc can only take on values of 0 or 1. 

We can only compare the rows directly to the intercept. So knowing a fish is a Parkki means its weight will be significantly less than a bream and so on.


Next let's look at Name and Length together. We get a much better fit!
```{r}
model_name_ln <- lm (Weight_g ~ Name + Length_cm, fish)
summary(model_name_ln)
```


Finally, we can center the Length which makes the intercept more interpretable. It will tell us now what the expected weight of a bream is that is of the same length as the mean of all the fist.
```{r}
model_name_lnc <- lm (Weight_g ~ Name + LengthCenter, fish)
summary(model_name_lnc)
```
In this linear equation, if LengthCenter is 10cm it means we are considering fish that 10cmn longer than the mean of the length of all the fish.


Here we can visualize the line in 7 dimensional space projected by fish type onto 2 dimensions...
```{r}
ggplot(data=fish, aes(x=LengthCenter, y=Weight_g,color=Name)) + geom_point() + geom_smooth(method='lm') 
```



