---
title: "R Notebook on Regression (based on Winter 2020)"
output: html_notebook
---

First we load the tidyverse and read the "poverty.csv" as a tibble.

```{r}
library(broom)
library(tidyverse)
poverty <- read_csv('poverty.csv')
```
Here are the variables we will use in our regression analysis.
poverty$PovPct
poverty$Brth15to17
poverty$Brth18to19
poverty$ViolCrime
poverty$TeenBrth

Next we make our linear model, which essentially just finds the coefficients that minimize the residuals.
```{r}
model <- lm(poverty$Brth15to17 ~ poverty$PovPct)
coef(model)
summary(model)
```

Finally, let's plot the model using Winter 2020's plotting recommendations.
```{r}
poverty %>% ggplot(aes(x = poverty$PovPct, y = poverty$Brth15to17)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()
```

That's all review of what we did before. 

Next let's try to predict birth rate given poverty level and violent crime level.
```{r}
model <- lm(poverty$Brth15to17 ~ poverty$PovPct + poverty$ViolCrime)
coef(model)
summary(model)
```
let's examine this using the tidy package.
```{r}
tidy(model) %>% select(term, estimate)
```

We can round the values as follows.

```{r}
tidy(model) %>% select(term, estimate) %>% mutate(estimate = round(estimate,2))
```
This means the predicted birth rate for ages 15-17 is given by the equation 

Brth15to17 = 5.98 + 1.04PovPct + 0.34ViolCrime

Let's standardize the predictors.
```{r}
poverty <- mutate(poverty, PovPct_z = scale(PovPct), ViolCrime_z = scale(ViolCrime))
tibble(poverty)
```

So we make a standardized model.
```{r}
model_z <- lm(poverty$Brth15to17 ~ poverty$PovPct_z + poverty$ViolCrime_z)
tidy(model_z) %>% select(term, estimate) %>% mutate(estimate = round(estimate,2))
```
This means for each increase in PovPct by 1 standard deviation (holding other variables constant) birth rate increases by 4.43. Similarly, for each increase in ViolCrime by 1 standard deviation (holding other variables constant) birth rate increases by 3.07. This implies that poverty has a larger impact on birth rate of 15-17 year olds than violent crime.


```{r}
res <- residuals(model_z)
plot(fitted(model_z),res)
```

```{r}
cor(poverty$PovPct,poverty$ViolCrime)
cor(poverty$PovPct,poverty$Brth15to17)
cor(poverty$Brth15to17,poverty$ViolCrime)
```
Correlation is not causation!!
