---
title: "R Notebook on Regression (based on Winter 2020)"
output: html_notebook
---

First we load the tidyverse and read the "poverty.csv" as a tibble.

```{r}
library(tidyverse)
poverty <- read_csv('poverty.csv')
```


Next we identify the x and y variables we will use in our regression analysis.
```{r}
x <- poverty$PovPct
y <- poverty$Brth15to17
# y <- poverty$Brth18to19
# y <- poverty$ViolCrime
# y <- poverty$TeenBrth
```

Let's plot them!

```{r}
plot(x, y, pch=19)
```
Next we make our linear model, which essentially just finds the coefficients that minimize the residuals.
```{r}
model <- lm(y ~x)
coef(model)
```
This means that the equation it found is y = ax + b with a being the value under x and b being the intercept.
So we can check to see what we might expect if the poverty rate is 15% like this.
```{r}
a <- coef(model)[2]
b <- coef(model)[1] 
a*15 + b
```
We can ignore the 'x' there -- that is just a leftover label from the coef data frame.

The summary function summarizes key aspects of the model.
```{r}
summary(model)
```


The broom library provides more visually appealing ways to examine your data.

```{r}
library(broom)
tidy(model)
```
```{r}
glance(model)
```

Finally, let's plot the model using Winter 2020's plotting recommendations.
```{r}
poverty %>% ggplot(aes(x = x, y = y)) +
  geom_point() + geom_smooth(method = 'lm') +
  theme_minimal()
```

